Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-13 joint_inf.tex
add eqn number to block
add normalizing factor and verify x/y are maxima

@@

    &\qquad + p_{13}\cdot\log(1-w'),

@@

    &\qquad + p_{13}\cdot\log(1-w').

@@

a_{1} = 3-p_{1}-p_{3}-p_{12}-p_{23},
a_{2} = 2(p_{1}+p_{3}+p_{12}+p_{23}),
a_{3} = 3-p_{2}-p_{12}-p_{23}-p_{123},
a_{4} = 2(p_{2}+p_{12}+p_{23}+p_{123}),

@@

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then, to bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

The full likelihood in this case is then bounded below as

@@

    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\

@@

    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')

@@

    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})),

@@

i.e.\ the maximum is achieved at $\{1,1-b,1-2p_{\emptyset}\}$.

@@

    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).

>>>>

    &\qquad - \log(8) \\

@@

    &\qquad + p_{13}\cdot\log(1-w') \\
    &\qquad - \log(8),

@@

    &\qquad + p_{13}\cdot\log(1-w') \\
    &\qquad - \log(8).

@@

\begin{equation}
    \begin{aligned}
        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
    \end{aligned}
    \label{eq:a_const}
\end{equation}

@@

\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
we have
\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
The same approach works for $\hat{y}'$.
The upper bound for the likelihood is then maximized at

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
We also now need to check that
$$
\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob},
$$
p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
$$
We see that
$$
1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
$$
and
$$
y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
$$
the last inequality holding if $x \le 2$.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.

To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad - \log(8) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad - \log(8) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

The full likelihood in this case is then bounded below by

@@

    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\

@@

    &\qquad + \log(1-w')

@@

    &\qquad + b      \cdot\log(b)

@@

%das If we call p_{\emptyset} just p, we're maximizing
%   p log (1 + w) + (1 - p) log (1 - w).
%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
%So you're right. Good catch.
%
%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
i.e.\ the maximum is achieved at $\{1,1-b,0\}$.

@@

    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
